---
title: "Biomarkers of ASD"
subtitle: ""
author: "Sophie Shi, Keon Bruce Dibley, Colin Tien Nguyen, Pramukh Shubh Shankar"
date: last-modified
published-title: "Updated"
editor: visual
format: html
code-copy: true
execute:
  message: false
  warning: false
  echo: false
  cache: true
---

Use this as a template. Keep the headers and remove all other text. In all, your report can be quite short. When it is complete, render and then push changes to your team repository.

```{r}
# load any other packages and read data here
library(tidyverse)
library(ggplot2)

load("./data/biomarker-clean.RData")
```

## Abstract

Write a brief one-paragraph abstract that describes the contents of your write-up.

## Dataset

*Write a brief data description, including: how data were obtained; sample characteristics; variables measured; and data preprocessing. This can be largely based on the source paper and should not exceed 1-2 paragraphs.*

Serum samples from 76 boys with ASD and 78 typically developing (TD) boys, 18 months-8 years of age, were analyzed to identify possible early biological markers for ASD. The ASD group was comprised of 76 subjects with a mean age of 5.6 years (SD 1.7 years). The TD group was comprised of 78 subjects with a mean age of 5.7 years (SD 2.0 years). Measured variables include ADOS and 1317 proteins. For the ASD group, ADOS diagnostic algorithms consisting of two behavioral domains: Social Affect (SA) and Restricted and Repetitive Behaviors (RRB) were used to determine an ADOS total score, which provides a continuous measure of overall ASD symptom severity. All subjects were healthy, defined as being fever-free for 24 hours, and presenting with no clinical symptoms. A fasting blood draw was performed on ASD and TD subjects between the hours of 8--10 am in a 3.5 ml Serum Separation Tube using standard venipuncture technique. 

In the published data, 192 proteins failed to pass quality control (QC). After removing these proteins, 1,125 proteins were analyzed. The protein abundance data were normalized by taking log10 transform and then z-transformation. To deal with outliers, any z-transformed values less than -3 and greater than 3 were clipped to -3 and 3, respectively. 

## Summary of published analysis

*Summarize the methodology of the paper in 1-3 paragraphs. You need not explain the methods in depth as we did in class; just indicate what methods were used and how they were combined. If possible, include a diagram that depicts the methodological design. (Quarto has support for [GraphViz and Mermaid flowcharts](https://quarto.org/docs/authoring/diagrams.html).) Provide key results: the proteins selected for the classifier and the estimated accuracy.*

Hewitson et al. first used the R package 'randomForest' to train the RF models and calculate feature importance. They chose MeanDecreaseGini as the surrogate representing a protein's importance in predicting ASD versus TD. With normalized data, they trained an RF model 1000 times. Each protein's importance value was averaged over the 1000 runs. The 10 proteins with the highest averaged importance values were chosen for the RF-based prediction model. For multiple testing, Hewitson et al. used t-tests to select the 10 proteins with the most highly significant t-test values for the prediction model. Thirdly, a correlation approach was used to calculate each protein's correlation with ADOS total scores (SA + RRB) as a measure of ASD severity. Based upon the absolute values of each protein's correlation coefficient, the 10 most highly correlated proteins were selected as the correlation-based predictive proteins. After identifying the top-10 predictive proteins from each of the 3 models, they found 5 proteins that were common to each method used which are considered 'core' proteins. A logistic regression model was used with datasets based upon the RF model, the t-test model and the correlation model, taking the subjects' assigned group (ASD or TD) as output variables to evaluate the 13 additional proteins. 

Five core proteins mitogen-activated protein kinase 14 (MAPK14), immunoglobulin D (IgD), dermatopontin (DERM), ephrin type-B receptor 2 (EPHB2), and soluble urokinase-type plasminogen activator receptor (suPAR) are used. And four additional proteins are selected for their resulted increase in AUC when being added which are receptor tyrosine kinase-like orphan receptor 1 \[ROR1\], platelet receptor Gl24 \[GI24\], eukaryotic translation initiation factor 4H \[elF-4H\], and arylsulfatase B \[ARSB\]. Combining the 5 core proteins with the additional 4 proteins resulted in an AUC = 0.860±0.064, with a sensitivity = 0.833±0.118, and specificity = 0.846±0.118, and represents the 9 optimal proteins (AUC_Optimal).

## Findings

*Summarize your findings here. I've included some subheaders in a way that seems natural to me; you can structure this section however you like.*

### Impact of preprocessing and outliers

Tasks 1-2

```{r, message=FALSE}
#SETUP
# get names
var_names <- read_csv('data/biomarker-raw.csv', 
                      col_names = F, 
                      n_max = 2, 
                      col_select = -(1:2)) %>%
  t() %>%
  as_tibble() %>%
  rename(name = V1, 
         abbreviation = V2) %>%
  na.omit()

# function for trimming outliers
trim <- function(x, .at){
  x[abs(x) > .at] <- sign(x[abs(x) > .at])*.at
  return(x)
}

# read in data
biomarker_clean <- read_csv('data/biomarker-raw.csv', 
                            skip = 2,
                            col_select = -2L,
                            col_names = c('group', 
                                          'empty',
                                          pull(var_names, abbreviation),
                                          'ados'),
                            na = c('-', '')) %>%
  filter(!is.na(group)) %>%
  # log transform, center and scale, and trim
  mutate(across(.cols = -c(group, ados), 
                ~ trim(scale(log10(.x))[, 1], .at = 3))) %>%
  # reorder columns
  select(group, ados, everything())

##TASK 1
set.seed(333) # Set seed for reproducibility
sample_proteins <- sample(names(biomarker_clean)[-c(1, ncol(biomarker_clean))], 5)  # Exclude 'group' and 'ados'

# Pivot data to a long format for easier plotting
biomarker_long <- biomarker_clean %>%
  select(all_of(sample_proteins)) %>%
  pivot_longer(cols = everything(), names_to = "protein", values_to = "value")

# Plot histograms of raw values for the sampled proteins
ggplot(biomarker_long, aes(x = value)) +
  geom_histogram(bins = 30, color = "black", fill = "skyblue", alpha = 0.7) +
  facet_wrap(~ protein, scales = "free") +
  labs(title = "Distribution of Raw Protein Levels",
       x = "Protein Level",
       y = "Frequency") +
  theme_minimal()
```

The data seems to be heavily skewed to the right. By doing a log transform, the skewness of the data would be lessened, making the patterns easier to analyze as most models assume constant variance.

```{r}
## task 2
# Read variable names as before
var_names <- read_csv('data/biomarker-raw.csv', 
                      col_names = F, 
                      n_max = 2, 
                      col_select = -(1:2)) %>%
  t() %>%
  as_tibble() %>%
  rename(name = V1, 
         abbreviation = V2) %>%
  na.omit()

# Read in data without trimming
biomarker_clean_no_trim <- read_csv('data/biomarker-raw.csv', 
                                    skip = 2,
                                    col_select = -2L,
                                    col_names = c('group', 
                                                  'empty',
                                                  pull(var_names, abbreviation),
                                                  'ados'),
                                    na = c('-', '')) %>%
  filter(!is.na(group)) %>%
  # log transform, center and scale without trimming
  mutate(across(.cols = -c(group, ados), 
                ~ scale(log10(.x))[, 1])) %>%
  # reorder columns
  select(group, ados, everything())







# Define threshold for identifying outliers
outlier_threshold <- 3

# Count outliers per subject
outlier_counts <- biomarker_clean_no_trim %>%
  rowwise() %>%
  mutate(outlier_count = sum(abs(c_across(-c(group, ados))) > outlier_threshold, na.rm = TRUE)) %>%
  ungroup()

############ EDA

# Display the summary of outliers per group
outlier_summary <- outlier_counts %>%
  group_by(group) %>%
  summarize(mean_outliers = mean(outlier_count),
            max_outliers = max(outlier_count),
            subjects_with_outliers = sum(outlier_count > 0))

# See that the average amount of outliers for ASD is 13.2 and 17.6 for TD







biomarker_data_outliers <- read_csv('data/biomarker-raw.csv', 
                           skip = 2,
                           col_select = -2L,
                           col_names = c('group', 'empty', pull(var_names, abbreviation), 'ados'),
                           na = c('-', '')) %>%
  filter(!is.na(group)) %>%
  mutate(across(-c(group, ados), ~ abs(scale(log10(.x))[, 1]) > 3, .names = "outlier_{.col}"))









# Count outliers for each subject
outlier_summary_2 <- biomarker_data_outliers %>%
  mutate(total_outliers = rowSums(select(., starts_with("outlier_")))) %>%
  select(group, total_outliers)

# Display the table
outlier_summary_2
View(outlier_summary_2)


#Every subject had at least 1 outlier except for 2
#5 subjects had a dispropportionate amount of outliers (>100)
## 2 of these subjects were ASD and 4 were TD
#Outliers did not appear more frequent in one group in regards to another

# Create the scatter plot
ggplot(outlier_summary_2, aes(x = group, y = total_outliers, color = group)) +
  geom_jitter(width = 0.2, height = 0, size = 3, alpha = 0.7) +
  labs(
    title = "Distribution of Outliers by Group",
    x = "Group",
    y = "Total Outliers"
  ) +
  theme_minimal() +
  scale_color_manual(values = c("ASD" = "skyblue", "TD" = "salmon"))




```

We found that the average amount of outliers for ASD is 13.2 and 17.6 for TD. Also, we can see that very subject had at least 1 outlier except for 2. 5 subjects had a dispropportionate amount of outliers (\>100), and 2 of these subjects were ASD while 4 were TD. As you can see in our graph, outliers didn't appear more frequent in one group in regards to another.

### Methodlogical variations

Task 3

### Improved classifier

Task 4

```{r}
library(glmnet)
library(tidyverse)
library(tidymodels)
library(tibble)
library(modelr)


set.seed(123)


biomarker_LASSO = biomarker_clean %>% 
  select(-ados) %>% 
  mutate(class = as.numeric(group == 'ASD'))
partitions <- biomarker_LASSO %>%
  initial_split(prop = 0.8)

x_train <- training(partitions) %>%
  select(-group, -class) %>%
  as.matrix()
y_train <- training(partitions) %>%
  pull(class)

lambda_test <- cv.glmnet(x_train, y_train, family = 'binomial', nfolds=5)
lambda_min <- lambda_test$lambda.1se
lambda_min = exp(-1.75)
final_fit <- glmnet(x_train, y_train, family = 'binomial', lambda = lambda_min)
final_fit_df = tidy(final_fit)


proteins_panel = final_fit_df %>%
  filter(term != "(Intercept)") %>% 
  pull(term)
biomarker_panel <- biomarker_clean %>%
  select(group, any_of(proteins_panel)) %>%
  mutate(class = (group == 'ASD')) %>%
  select(-group)

biomarker_panel

biomarker_split <- biomarker_panel %>%
  initial_split(prop = 0.8)

fit <- glm(class ~ ., 
           data = training(biomarker_split), 
           family = 'binomial')

class_metrics <- metric_set(accuracy)

testing(biomarker_split) %>%
  add_predictions(fit, type = 'response') %>%
  mutate(est = as.factor(pred > 0.5), tr_c = as.factor(class)) %>%
  class_metrics(estimate = est,
                truth = tr_c, pred,
                event_level = 'second')



```

First, we split our data into training and testing sets, with 80% of the data being training data. Next, we performed cross validation on the training set to get a panel of proteins. Then, we fit a binomial generalized linear model to this panel of proteins and evaluated our metrics. We reported an accuracy of 0.70967 on our panel of proteins, which included "CXCL16, soluble", "MAPK2", "Calcineurin", "IgD", and "DERM".
